---
title: "MUR23 phytoplankton proxies (lipid-based and sedDNA)"
author: "Antonia Klatt"
date: "2026-05-27"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)

#Load packages

library(readr)
library(ggplot2)
library(gridExtra)
library(grid)
library(cowplot)
library(reshape2)
library(patchwork)
library(caret)
library(gam)
library(vegan)
library(dplyr)
library(rioja)
library(ggdendro)
library(stringr)
library(tidyr)

#Import data

#age-depth model

MUR23_rplum <- read.csv (here("01_Data","Murtensee","MUR23_D1_91_ages_run_260526.csv"), header = TRUE , sep = "," , stringsAsFactors = FALSE, na.strings = "#NV")

#select every second row of MUR23_rplum --> specific subsampling depths

MUR23_age_depth <- MUR23_rplum[3:87,] %>% filter(row_number() %% 3 == 1)


#Lipid data

##Phytol:sterol index (PSI)

MUR23_D1_phytol.sterol.ratio <- read.csv(here("01_Data","Murtensee","MUR23_D1_phytol.sterol.ratio.csv"), row.names=1, header = TRUE , sep = ";" , stringsAsFactors = FALSE, na.strings = "#NV")

MUR23_D1_phytol.sterol.ratio$mean_age <- MUR23_age_depth$mean


##absolute alcohol concentrations (ug/g TOC)

MUR23_D1_alcohol_conc.a <- read.csv(here("01_Data","Murtensee","MUR23_D1_F3_ug_gTOC.csv"),header = TRUE , sep = ";" , stringsAsFactors = FALSE, na.strings = "#NV")

#remove random NAs from import

MUR23_D1_alcohol_conc <- MUR23_D1_alcohol_conc.a[1:29,]

MUR23_D1_alcohol_conc$mean_age <- MUR23_age_depth$mean


##relative alcohol concentrations (%)

MUR23_alcohols_rel_conc <- read.csv(here("01_Data","Murtensee","MUR23_D1_F3_rel_conc_woAlkanols.csv"), row.names=1, header = TRUE , sep = ";" , stringsAsFactors = FALSE, na.strings = "#NV")

row.names(MUR23_alcohols_rel_conc) <- MUR23_age_depth$mean


##d2H C16:0 Acid/Phytol

MUR23_D1_d2H_C16.0_phytol <- read.csv(here("01_Data","Murtensee","MUR23_d2H_C16.0_phytol.csv"), row.names=1, header = TRUE , sep = ";" , stringsAsFactors = FALSE, na.strings = "#NV")

MUR23_D1_d2H_C16.0_phytol$mean_age <- MUR23_age_depth$mean


#23S rRNA amplicon sequenceing variants (ASV) data (sedDNA)

#ASV count data

Swiss_Lakes_ASV_count <- read.delim(here("Microbiota_Data", "05_OUTPUT",'COUNT_phytoplankton.tab'))

#ASV taxonomy data (NCBI)

Swiss_Lakes_ASV_tax <- read.delim(here("Microbiota_Data", "05_OUTPUT",'TAXA_phytoplankton.tab'))


#extract ASV counts from Murtensee

#taxonomic phylum level

MUR_ASV_count.a <- str_detect(names(Swiss_Lakes_ASV_count), "MUR23")

MUR_ASV_count <- Swiss_Lakes_ASV_count[, MUR_ASV_count.a]

MUR_ASV_count$phylum <- Swiss_Lakes_ASV_tax$phylum

```

## Stratigraphically constrained clustering (CONISS) of relative alcohol concentrations

```{r}

#create dissimilarity matrix based on Bray Curtis distance (like in NDMS)

MUR23_alc_dist <- vegdist(MUR23_alcohols_rel_conc, method="bray", binary=FALSE, diag=FALSE, upper=FALSE)


#CONISS constrained cluster analysis chclust(d, method = "coniss")

MUR23_alc_CONISS <-chclust(MUR23_alc_dist, method = "coniss")


plot(MUR23_alc_CONISS, cex.axis=0.5, cex= 0.75, main= "CONISS MUR23 Alcohols")

bstick(MUR23_alc_CONISS, ng=10, plot=TRUE, title = "Broken stick model CONISS MUR23 alcohols") #3 clusters significant


#extract cluster number from chclust

MUR23_alc_groups<-cutree(MUR23_alc_CONISS, k=3) #k=3 since we have 3 significant clusters according to bstick

MUR23_alc_clust_Age <-cbind.data.frame(MUR23_age_depth$mean, MUR23_alc_groups)

colnames(MUR23_alc_clust_Age) <- c("Year", "MUR23_CONISS_cluster")


#group1 :2019.2-2007.2
#group2: 2001.0-1939.8
#group3: 1933.7-1854.6

#draw lines between edge values from zones in ggplot
#e.g. (younger value group2 - ((younger group2-older value group1)/2)
#--> 1936.85; 2004.1


#extract dendrogram

MUR23_alc_dendro <- as.dendrogram(MUR23_alc_CONISS)

MUR23_dend_data <- dendro_data(MUR23_alc_dendro)

#replace labels (order) in dendrogram by actual ages with approxfun for aligning plots

#Code adapted from https://stackoverflow.com/questions/65207033/how-do-i-plot-dendrogram-leaves-by-a-given-value-depth-in-a-sediment-core-rath (accessed 22.07.2025)

dendAge <- approxfun(MUR23_dend_data$labels$x, as.numeric(as.character(MUR23_dend_data$labels$label))) 
MUR23_dend_data$segments$x <- dendAge(MUR23_dend_data$segments$x)
MUR23_dend_data$segments$xend <- dendAge(MUR23_dend_data$segments$xend)


pdend.MUR <- ggplot(segment(MUR23_dend_data)) +
  annotate("rect", xmin =  1950 ,xmax = 1980 ,ymin = -Inf, ymax = Inf, alpha=0.065, fill="red1")+ 
  ggtitle(expression('CONISS'['Alcohols']), subtitle="CONISS")+
  geom_vline(xintercept=1936.85, color = "lightsteelblue4", linetype="dashed", linewidth=0.75)+
  geom_vline(xintercept=2004.1, color = "lightsteelblue4", linetype="dashed", linewidth=0.75)+
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  scale_x_continuous(limits= c(1854,2023), expand=c(0,0))+
  scale_y_continuous()+
  theme_dendro()+
  theme( plot.title = element_text(size=37.5, hjust = 0.5),
         plot.subtitle = element_blank(),
         plot.margin = unit(c(0.5,0.85,0,0.5), "cm"))

```

## create plot with CONISS cluster dendrogram of alcohol concentrations, PSI values, d2H C16:0 Acid/Phytol values, relative abundance of Cyanobacteriota (23S rRNA ASVs)

```{r}

#LOESS fit lipid-based proxies

#according to https://www.statology.org/loess-regression-in-r/ (29.07.2025) and https://www.statology.org/leave-one-out-cross-validation-in-r/ (11.08.2025)

#define k-fold cross validation method with leave-one-out cross-validation (LOOCV)

#perform cross-validation using smoothing spans ranging from 0.1 to 0.9

ctrl <- trainControl(method = "LOOCV")
grid <- expand.grid(span = seq(0.1, 0.9, len = 9), degree = 1)

#LOESS fit PSI

MUR_PSI_model_cv <- train(phytol.sterol.ratio.wo.4a_17.20_chol ~ mean_age, data = MUR23_D1_phytol.sterol.ratio, 
                          method = "gamLoess", tuneGrid=grid, trControl = ctrl)

#print results of k-fold cross-validation

print(MUR_PSI_model_cv) #The final values used for the model were span = 0.4 and degree = 1.

#final model with span = 0.5 

MUR_PSI_loess.4 <- loess(phytol.sterol.ratio.wo.4a_17.20_chol ~ mean_age, data = MUR23_D1_phytol.sterol.ratio, span=.4)

MUR_PSI_smooth.a <- as.data.frame(predict(MUR_PSI_loess.4, se = TRUE))

#calculate 95% confidence interval
#https://stackoverflow.com/questions/22717930/how-to-get-the-confidence-intervals-for-lowess-fit-using-r (01.08.2025)

MUR_PSI_smooth.a$upper <- MUR_PSI_smooth.a$fit + qt(0.975,MUR_PSI_smooth.a$df)*MUR_PSI_smooth.a$se

MUR_PSI_smooth.a$lower <- MUR_PSI_smooth.a$fit - qt(0.975,MUR_PSI_smooth.a$df)*MUR_PSI_smooth.a$se

MUR_PSI_smooth_wAge <- cbind.data.frame(MUR23_D1_phytol.sterol.ratio, MUR_PSI_smooth.a)


MUR23_PhySter <- ggplot() +
  geom_ribbon(data=MUR_PSI_smooth_wAge, aes(x= mean_age, ymax = upper, ymin = lower), fill = "cyan3", 
              alpha = 0.175) + 
  geom_line(data = MUR_PSI_smooth_wAge[!is.na(MUR_PSI_smooth_wAge$phytol.sterol.ratio.wo.4a_17.20_chol),] , 
            aes(x = mean_age, y = phytol.sterol.ratio.wo.4a_17.20_chol), color= 'darkslategray', linewidth= 1) +  
  scale_x_continuous(limits= c(1854,2023), expand=c(0,0))+
  scale_y_continuous(position="left", limits=c(0,0.8))+
  xlab("Year CE") + ylab(expression(atop("phytol:sterol index")))+ 
  theme(axis.line = element_line(colour = "black"), 
        panel.background = element_rect(fill = NA, color = NA),
        plot.background  = element_rect(fill = NA, color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(size=37.5,color="black"),
        #axis.text.y = element_blank(),
        axis.title.y = element_text(size=37.5,color="black"),
        axis.title.x = element_blank(),
        axis.line.x =element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position='none',
        plot.margin = unit(c(0,0.85,0.25,0.5), "cm"))


#LOESS fit d2H C16:0/phytol

#EXCLUDE CONTAMINANT SAMPLES (d2H phytol > -300 permill)

MUR23_D1_d2H_C16.0_phytol_wocont <- MUR23_D1_d2H_C16.0_phytol[-which(MUR23_D1_d2H_C16.0_phytol$phytol_d2H.VSMOW > -300),] 


MUR_d2H_model_cv <- train(epsilon_C16.0_phytol ~ mean_age, data = na.omit(MUR23_D1_d2H_C16.0_phytol_wocont), 
                          method = "gamLoess", tuneGrid=grid, trControl = ctrl)

#print results of k-fold cross-validation

print(MUR_d2H_model_cv) #The final values used for the model were span = 0.5 and degree = 1.

#final model with span = 0.5 

MUR_d2H_loess.5 <- loess(epsilon_C16.0_phytol ~ mean_age, data = na.omit(MUR23_D1_d2H_C16.0_phytol_wocont), span=.5)

MUR_d2H_smooth.a <- as.data.frame(predict(MUR_d2H_loess.5, se = TRUE))

#calculate 95% confidence interval
#https://stackoverflow.com/questions/22717930/how-to-get-the-confidence-intervals-for-lowess-fit-using-r (01.08.2025)

MUR_d2H_smooth.a$upper <- MUR_d2H_smooth.a$fit + qt(0.975,MUR_d2H_smooth.a$df)*MUR_d2H_smooth.a$se

MUR_d2H_smooth.a$lower <- MUR_d2H_smooth.a$fit - qt(0.975,MUR_d2H_smooth.a$df)*MUR_d2H_smooth.a$se

MUR_d2H_smooth_wAge <- cbind.data.frame(na.omit(MUR23_D1_d2H_C16.0_phytol_wocont), MUR_d2H_smooth.a)


MUR23_d2H <- ggplot() +
  geom_ribbon(data=MUR_d2H_smooth_wAge, aes(x= mean_age, ymax = upper, ymin = lower), fill = "cyan3", 
              alpha = 0.175) + 
  geom_line(data = MUR_d2H_smooth_wAge[!is.na(MUR_d2H_smooth_wAge$epsilon_C16.0_phytol),] , 
            aes(x = mean_age, y = epsilon_C16.0_phytol), color= 'darkslategray', linewidth= 1) +  
  scale_x_continuous(limits= c(1854,2023), expand=c(0,0))+
  scale_y_continuous(position="right", limits=c(115,305))+
  xlab(expression(atop("","Year CE"))) +  ylab(expression(atop(delta^2*H['C16:0 Acid/Phytol']~("\u2030"), "")))+
  theme(axis.line = element_line(colour = "black"), 
        panel.background = element_rect(fill = NA, color = NA),
        plot.background  = element_rect(fill = NA, color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(size=37.5,color="black"),
        #axis.text.y = element_blank(),
        axis.title.y = element_text(size=37.5,color="black"),
        axis.title.x = element_blank(),
        axis.line.x =element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position='none',
        plot.margin = unit(c(0.25,0.85,0.25,0.5), "cm"))


#relative abundance Cyanobacteriota (23S rRNA ASVs)

#sum rows with same phylum  

MUR_ASV_count_phylum <- aggregate(.~phylum,MUR_ASV_count,FUN=sum)

#replace colnames with age

colnames(MUR_ASV_count_phylum)

#select specific subsampling depths of DNA (same as lipids)

MUR23_DNA_age <- filter(MUR23_rplum, depth %in% c(11,14,23,29,32,35,38,44,47,53,65,71,77))

#round to full numbers 

MUR23_DNA_age2 <- round(MUR23_DNA_age$mean, digits=0)

#replace sample names with age

colnames(MUR_ASV_count_phylum) <- c("phylum", MUR23_DNA_age2)

MUR_ASV_count_phylum.rev <- rev(MUR_ASV_count_phylum)

#extract Cyanobacteriota

MUR_ASV_count_cyano <- subset(MUR_ASV_count_phylum.rev, grepl('Cyanobacteriota', phylum))

MUR_ASV_count_cyano2 <- as.data.frame(t(MUR_ASV_count_cyano[,1:13]))

MUR_ASV_count_cyano_age_CE <- as.numeric(rownames(MUR_ASV_count_cyano2))

MUR_ASV_count_cyano_rel <- as.numeric(MUR_ASV_count_cyano2[,1:1])

MUR_ASV_count_cyano3 <- cbind.data.frame(MUR_ASV_count_cyano_age_CE, MUR_ASV_count_cyano_rel)

colnames(MUR_ASV_count_cyano3) <- c( "age_CE", "Cyanobacteriota")

pMUR23_DNA_Cyano <- ggplot() +
  geom_area(data = MUR_ASV_count_cyano3 , aes(x = age_CE, y =Cyanobacteriota), fill="darkslategray", 
            alpha=0.75)+
  scale_x_continuous(limits= c(1854,2023), breaks=c(1860,1880,1900,1920,1940,1960,1980,2000,2020),expand=c(0,0))+
  scale_y_continuous(position="left", limits=c(0,105), breaks=c(25,50,75,100), expand= c(0,0))+
  xlab(expression(atop("","Year CE"))) + ylab(expression(atop("% Cyanobacteriota")))+ 
  theme(axis.line = element_line(colour = "black"), 
        panel.background = element_rect(fill = NA, color = NA),
        plot.background  = element_rect(fill = NA, color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(size=37.5,color="black", angle=45, vjust=0.5),
        axis.text.y = element_text(size=37.5,color="black"),
        axis.title.x = element_text(size=37.5,color="black"),
        axis.title.y = element_text(size=37.5,color="black"),
        legend.text=element_text(size=37.5, margin=margin(l=20)),
        legend.title=element_text(size=37.5, margin=margin(b=20)),
        legend.key=element_blank(),
        legend.key.spacing.x = unit(1, 'cm'),
        legend.position = "bottom",
        plot.margin = unit(c(0,0.65,0.5,0.5), "cm"))

Fig4a <- plot_grid(pdend.MUR, MUR23_PhySter, MUR23_d2H, pMUR23_DNA_Cyano,  
          nrow=4,rel_heights= c(0.5, 1.05,1.075, 1.425), 
          axis= "rl", align = "v")

```

# Fig. 4a

```{r, fig.width = 16.5, fig.height = 23.5, echo=FALSE}

plot(Fig4a)

```

## create plot with CONISS cluster dendrogram of alcohol concentrations and absolute concentrations of dinostanol and dinosterol

```{r}

MUR23_dino <- ggplot() +
  annotate("rect", xmin =  1950 ,xmax = 1980 ,ymin = -Inf, ymax = Inf, alpha=0.065, fill="red1")+ 
  #geom_ribbon(data=GRE_dino_smooth_wAge, aes(x= year, ymax = upper, ymin = lower), fill = "plum1", alpha = 0.25) + 
  geom_line(data = MUR23_D1_alcohol_conc, 
            aes(x = mean_age, y =  Dinostanol_ug.gTOC), color= '#ccccff', linewidth=1.5) +
  geom_line(data = MUR23_D1_alcohol_conc, 
            aes(x = mean_age, y =  Dinosterol_ug.gTOC), color= '#660066', linewidth=1.5) +
  #reverse depth so it starts at zero
  #put the y axis labes on the opposite side so when its flipped it will appear at top
  geom_vline(xintercept=1936.85, color = "lightsteelblue4", linetype="dashed", linewidth=0.75)+
  geom_vline(xintercept=2004.1, color = "lightsteelblue4", linetype="dashed", linewidth=0.75)+
  #reverse depth so it starts at zero
  scale_x_continuous(limits= c(1854,2023), breaks=c(1860,1880,1900,1920,1940,1960,1980,2000,2020),expand=c(0,0))+
  scale_y_continuous(position="left", limits=c(-10,410),expand= c(0,0))+
  #, limits=c(-0.15,1.15), breaks = seq(0, 1, by = 1))+
  #this is how you reverse the look and order or the coordinates for the graph
  xlab(expression(atop("","Year CE"))) +  ylab(expression(atop(mu*g~'/g'['TOC'])))+
  theme(axis.line = element_line(colour = "black"), 
        panel.background = element_blank(),
        axis.text.x = element_text(size=37.5,color="black", angle=45, vjust=0.5),
        axis.text.y = element_text(size=37.5,color="black"),
        axis.title.x = element_text(size=37.5,color="black"),
        axis.title.y = element_text(size=37.5,color="black"),
        legend.text=element_text(size=37.5, margin=margin(l=20)),
        legend.title=element_text(size=37.5, margin=margin(b=20)),
        legend.key=element_blank(),
        legend.key.spacing.x = unit(1, 'cm'),
        legend.position = "bottom",
        plot.margin = unit(c(0,0.65,0.5,0.5), "cm"))

Fig4b <- plot_grid(pdend.MUR, MUR23_dino, 
          nrow=2,rel_heights= c(0.5, 1.41), 
          axis= "rl", align = "v")

```

# Fig. 4b

```{r, fig.width = 14.25, fig.height = 11.15, echo=FALSE}

plot(Fig4b)

```
