---
title: "Microbiome"
author: "Jan Wälchli"
date: "`r Sys.Date()`"
toc: true
toc-depth: 3
toc-location: left
fig-align: center
tbl-align: center
cap-location: top
lof: false
lot: false
format: pdf
execute: 
  echo: false
  message: false
  warning: false
  error: false
header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H} #keeps table in float, otherwise placed at top
---

```{r setup}

##clear the object from memory
rm(list=ls())

#global knitter option
knitr::opts_chunk$set(out.width = "90%")

# load packages
library(this.path)
library(tidyverse)
library(phyloseq)
library(vegan)
library(readxl)
library(cowplot)
library(ggh4x)
library(RColorBrewer)
library(gt)
library(dada2)

# set path to source file location
setwd(this.path::here())

# set global option for ggplot2
theme_set(theme_bw(base_size = 14))

#set seed
set.seed(100)

#import paths
paths <- list(bASV_table = c("../02_dadapipe/10_OUTPUT/bacteria_COUNT100.tab"),
              TAXA = c("../03_NCBI_taxonomy/bacteria_TAXA100_NCBI.tab"),
              SEQ =c("../02_dadapipe/10_OUTPUT/bacteria_SEQ100.fasta")
              )

```

## Background

Add your text here

```{r COUNT import}

#import ASV table
COUNT <- read.delim(paths$bASV_table, header=T, row.names=1, sep="\t")
COUNT <- t(COUNT)

# simplify names
colnames(COUNT) <- gsub("2504461930_", "", colnames(COUNT))
rownames(COUNT) <- gsub("OTU", "ASV", rownames(COUNT))

```

```{r TAXA import}

#import taxonomy table
TAXA <- read.table(paths$TAXA, row.names=1, sep="\t", blank.lines.skip = FALSE)

#rename
colnames(TAXA) <- c("kingdom", "phylum", "class", "order", "family", "genus", "species")
rownames(TAXA) <- gsub("OTU", "ASV", rownames(TAXA))
TAXA[is.na(TAXA)] <- "unassigned"
TAXA <- TAXA[rownames(COUNT),] #keep only ASVs which are in COUNT

#add ASV_ID to TAXAonomy file
TAXA$ASV_ID <- rownames(TAXA)

#levels
TAXA$phylum <- as.factor(TAXA$phylum)
levels(TAXA$phylum)[levels(TAXA$phylum) < 2 ] <- "unassigned"

```

```{r create design}

DESIGN <- data.frame(ID=colnames(COUNT)) %>%
  mutate(
    Lake = str_sub(ID, 1, 3),
    Year = str_sub(ID, 4, 5),
    Type = ifelse(str_detect(ID, regex("blank", ignore_case = TRUE)), "blank", "DNA")
  )

#define lake colors
lake_colors <- c("#D3BA68FF", "#D5695DFF", "#5D8CA8FF", "#65A479FF")
names(lake_colors) <- unique(DESIGN$Lake)

```

\pagebreak

## Description all data

### Sample Removal

```{r seq-numbers}

## define threshold 
b_rare <- 390000

samples_to_remove <- names(which(colSums(COUNT) < b_rare))
print(paste(length(samples_to_remove), "samples removed due to low number of sequences (threshold =", b_rare, "sequences)"))
print(samples_to_remove)

DESIGN <- DESIGN[!(DESIGN$ID %in% samples_to_remove),]
COUNT <- COUNT[,!(colnames(COUNT) %in% samples_to_remove)]
ASVs_to_remove <- names(which(rowSums(COUNT) == 0))
COUNT <- COUNT[!(rownames(COUNT) %in% ASVs_to_remove),]
TAXA <- TAXA[!(rownames(TAXA) %in% ASVs_to_remove),]

```

### Summary

We show the sum, range and median of sequences over all samples.

```{r seq-numbers summary}

#seq numbers
df <- data.frame(sum=NA, min=NA, max=NA, median=NA, ASVs=NA)
df$sum <- sum(colSums(COUNT))
df$min <- min(colSums(COUNT))
df$max <- max(colSums(COUNT))
df$median <- median(colSums(COUNT))
df$ASVs <- nrow(COUNT)

#out
#pander(df)
df %>%
  gt() %>%
  tab_style(style = cell_text(align="center"), locations = cells_body()) %>%
  tab_style(style = list(cell_text(weight = "bold", align = "center"), cell_fill("grey96")), locations = list(cells_column_labels())) %>%
  fmt_number(sep_mark = "'", decimals=0) %>%
  tab_options(table.width = px(250))

rm(df, ASVs_to_remove, samples_to_remove)

```

\pagebreak

### Sequencing depth plot

```{r seq depth plot}
#| fig-width: 6
#| fig-asp: 1 #width to height aspect ratio

#functions for plots
seq_nr_plot <- function(df, title="", ylim){
  seq_nr <- ggplot(data=df, aes(x=Lake, y=col_sum, fill=Lake, shape=Year)) + 
              geom_boxplot(position=position_dodge2(width=0.75, preserve="single"),outlier.colour = NA)+
              geom_jitter(aes(group=Lake), size=3, position=position_jitterdodge(jitter.width=0.2,dodge.width = 0.75))+
              ggh4x::facet_nested( ~ Lake, scale="free_x", space="fixed", nest_line = element_line())+ 
              geom_segment(x=-Inf,xend=Inf,y=(ylim*1.05),yend=(ylim*1.05))+ #add lines under WT / bx1
              geom_segment(x=-Inf,xend=Inf,y=0,yend=0,lwd=0.2, col="black")+ #box around grids
              theme_minimal()+
              ylim(0,ylim)+
              theme(legend.position="bottom", axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
                     panel.grid.minor = element_blank())+
              ylab("seq / samples") +
              scale_fill_manual(values=lake_colors) +
              scale_shape_manual(values=c("23"=21,"24"=24))
              #ggtitle(title)
  return(seq_nr)
}

samples_plot <- function(df){
  samples <-  ggplot(df, aes(x=x, y=col_sum, fill=Lake))+
                geom_bar(aes(width=1), size=0.1, stat = "identity", col="black", lty=0.8)+
                scale_fill_manual(values=lake_colors) +
                xlab("samples")+
                ylab("seq / samples")+
                ggh4x::facet_nested( ~ Lake, scale="free_x", space="fixed", nest_line = element_line())+ 
                geom_segment(x=-Inf,xend=Inf,y=(ylim*1.05),yend=(ylim*1.05))+ #add lines under WT / bx1
                geom_segment(x=-Inf,xend=Inf,y=0,yend=0,lwd=0.2, col="black")+ #box around grids
                theme_minimal()+
                theme(legend.position="bottom", axis.text.x=element_blank(), axis.ticks.x=element_blank(),panel.grid.minor = element_blank())
  return(samples)
}

#function to add x values for the barplot
add_x_values <- function(df, cols_to_order){
  df <- df[order(df$col_sum, decreasing = T),]
  for(o in cols_to_order){df <- df[order(df[,colnames(df)==o]),]} #sort
  groups <- do.call(paste0, c(as.list(df[,colnames(df) %in% cols_to_order]))) #groups
  table <- groups %>% table() #n samples per group
  table <- table[match(unique(groups), names(table))] #sort
  df$x <- table %>% as.vector() %>% lapply(., function(x) seq(1,x)) %>% unlist() #number string per group
  return(df)
}

#Bacteria
#boxplot
df <- data.frame(col_sum=colSums(COUNT),colnames=colnames(COUNT), Lake=DESIGN$Lake, Year=DESIGN$Year)
ylim=max(colSums(COUNT))
bac_seq_nr <- seq_nr_plot(df, "Bacteria: Sequencing depth", ylim) #create plot

## barplot
df <- add_x_values(df, c("Year", "Lake"))
bac_samples <- samples_plot(df) #create plot

#combine the plots (library cowplot)
plot <- plot_grid(bac_seq_nr + theme(legend.position="none"),
               bac_samples + theme(legend.position="none") + labs(colour=""),
               align = "hv",
               rel_heights = c(9,5),
               nrow=2)

legend <- get_legend(bac_seq_nr + theme(legend.position="bottom") + labs(shape="Field", fill="Conditioning"))
plot_grid(plot, legend, nrow=2, rel_heights=c(1, .1))

rm(bac_seq_nr, bac_samples, plot, legend, df, ylim)

```

\pagebreak

### Rarefaction plot

We plot a rarefaction plot to check if the sequence depth is enough to capture the microbial diversity.

```{r rarefaction}
#| fig-width: 6
#| fig-asp: 0.5 #width to height aspect ratio

#function to create rarefaction table
rarefaction_table <- function(count, design, name, step_size, sample_size){
  #rarefy
  pdf("/dev/null") #prevents drawing of standard rarefaction plot (we will create our own with colors)
  rarefaction <- vegan::rarecurve(t(count), step=step_size, sample_size, label = FALSE)
  names <- unlist(design[,colnames(design)==name])
  names(rarefaction) <- names
  invisible(dev.off())
  
  # long-transformation.
  protox <- mapply(FUN = function(x, y) {
    mydf <- as.data.frame(x)
    colnames(mydf) <- "value"
    mydf$species <- y
    mydf$subsample <- attr(x, "Subsample")
    mydf
  }, x = rarefaction, y = as.list(names(rarefaction)), SIMPLIFY = FALSE)
  rarefaction_long <- do.call(rbind, protox)
  rownames(rarefaction_long) <- NULL  # pretty

  #add soil, sampletype and conditioning
  rarefaction_long$Lake <- NA
  rarefaction_long$Year <- NA
  for (i in 1:nrow(rarefaction_long)) {
    rarefaction_long$Lake[i] <- as.character(design$Lake[names == rarefaction_long$species[i]])
    rarefaction_long$Year[i] <- as.character(design$Year[names == rarefaction_long$species[i]])
  }
  
  #rename
  colnames(rarefaction_long)[colnames(rarefaction_long)=="species"] <- "sample"

  #return
  return(rarefaction_long)
}

#function to create rarefaction plot
rarefaction_plot <- function(rarefaction_table, title="", rare_treshold, xlim){
   #limit for y axis
  ylim=max(rarefaction_table$value)
  #factor order
  #rarefaction_table$sampletype <- factor(rarefaction_table$sampletype, levels=c("root", "rhizo", "soil"))
  #rarefaction plot
  rareplot <- ggplot(rarefaction_table, aes(x = subsample, y = value, group = sample)) +
                    theme_bw() +
                    geom_line(aes(color=Lake))+
                    scale_color_manual(values = lake_colors)+
                    geom_vline(xintercept = rare_treshold, col="red", lwd=0.7, lty=3)+
                    ggh4x::facet_nested( ~ Lake, scale="fixed", space="free_x", nest_line = element_line())+ 
                    geom_segment(x=-Inf,xend=Inf,y=(ylim*1.05),yend=(ylim*1.05))+ #add lines under WT / bx1
                    geom_segment(x=-Inf,xend=Inf,y=0,yend=0,lwd=0.2, col="black")+ #box around grids
                    theme_minimal()+
                    ylim(0,ylim)+
                    xlim(0,xlim)+
                    theme(legend.position="bottom", panel.grid.minor = element_blank(), axis.text.x=element_text(angle=90))+
                    xlab("Sequencing Depth")+
                    ylab("Number of ASVs")
                    #ggtitle(title)
  return(rareplot)

}

#rarefaction table and plot
brarefaction <- rarefaction_table(COUNT, DESIGN, "ID", step_size=20000, sample_size=700000)
brarefaction_plot <- rarefaction_plot(brarefaction, "Bacteria: Rarefaction", NA, xlim=max(brarefaction$subsample))




#combine the two plots (library cowplot)
plot <- cowplot::plot_grid(brarefaction_plot + theme(legend.position="none"),
               align = "hv",
               nrow=1)

#add legend
legend <- cowplot::get_legend(brarefaction_plot + theme(legend.position="bottom") + labs(col="Conditioning"))
cowplot::plot_grid(plot, legend, nrow=2, rel_heights=c(1, .1))

```

## Normalization

### Kruskal-Wallis test

To decide on how to normalize the data we follow the recommendation of Weiss et al. (2017, Microbiome Journal) and inspect whether there are differences in sequencing depths between the different groups(soil-plant-conditioning) by using the non-parametric Kruskal-Wallis Test.

```{r seq numbers per group bacteria}

#BACTERIA
group <- as.factor(paste(DESIGN$Lake, sep="_"))
sample_depth <- colSums(COUNT)
COUNT_krusk <- kruskal.test(sample_depth ~ group)
COUNT_krusk

```
*Conclusion:* Since we find no significant difference between sample groups we normalise by total sum scaling

```{r COUNT norm}

#TSS normalisation
COUNT_norm <- apply(COUNT, 2, function(x) x *100 / sum(x)) #colSums(COUNT_norm) should be 100% per column

```

\pagebreak

### Outlier detection

We use the method CLOUD developed by Montassier et al. 2018, which is a non-parametric detection test for outliers. We perform the test with Bray-Curtis distances from the normalized data for each substrate and each plastic treatment individually. We set the number of nearest neighbors to 40% of the samples size and chose an empirical outlier percentile of 0.1. We remove all outliers from our data.

```{r outlier detection functions}

## OUTLIER DETECTION FUNCTION based on the CLOUD method (a non-parametric detection test) from Montassier et al. 2018
# d is a distance matrix
# returns piecewise distances of samples and their outlier percentile
# and a matrix of the repeated measures of distances
# k is number of neighbors choosen
# test.ix are the names of the tested samples
"piecewise_kn_V1" <- function(d, test.ix, k=X, ndim=-1){
  if(class(d) != 'matrix') d <- as.matrix(d)
  stats <- numeric(length(test.ix))
  pvals <- numeric(length(test.ix))
  for(i in 1:length(test.ix)){
    ref.ix <- test.ix[-i]
    keep.ix <- c(test.ix[i], ref.ix)
    if(ndim > -1){
      pc <- cmdscale(d[keep.ix,keep.ix,drop=F],k=ndim)
      d.i <- as.matrix(dist(pc))
    } else {
      d.i <- d[keep.ix,keep.ix,drop=F]
    }
    test.dist <- mean(sort(d.i[1,-1])[1:k])
    ref.dists <- numeric(length(ref.ix))
    for(j in 1:length(ref.ix)){
      ref.dists[j] <- mean(sort(d.i[-1,-1][j,-j]))
    }

    stats[i] <- test.dist / mean(ref.dists)
    pvals[i] <- mean(test.dist < ref.dists)
  }
  result <- list()
  result$stats <- stats
  result$pvals <- pvals
  outcome <- pvals <= 0.05
  result$lenght <- length(outcome[outcome==TRUE])
  return(result)
}


#function to detect outliers in our data
detect_outliers <- function(DESIGN, COUNT_norm, Lake){

  #subset
  samples_sub <- DESIGN$ID[DESIGN$Lake == Lake]
  COUNT_sub <- COUNT_norm[,colnames(COUNT_norm) %in% samples_sub]
  bdist <- vegan::vegdist(t(COUNT_sub), method="bray")

  #test
  k <- floor(ncol(COUNT_sub) * 0.4) #number of nearest neighbors, here we chose 40% of sample size
  test <- piecewise_kn_V1(d = bdist, test.ix = colnames(COUNT_sub), k = k) #CLOUD test from Montassier et al. 2018 (CLOUD function)
  outliers <- colnames(COUNT_sub)[test[["pvals"]]<=0.1] #outliers (p <= 0.1)

  return(outliers)

}

```

```{r outlier detection}

#get outliers
outliers_bac <- c(
  detect_outliers(DESIGN=DESIGN, COUNT_norm=COUNT_norm, Lake="GRE"),
  detect_outliers(DESIGN=DESIGN, COUNT_norm=COUNT_norm, Lake="MUR"),
  detect_outliers(DESIGN=DESIGN, COUNT_norm=COUNT_norm, Lake="NEU"),
  detect_outliers(DESIGN=DESIGN, COUNT_norm=COUNT_norm, Lake="ZUG")
)


#output
outliers <- data.frame(Species=c("Bacteria"),Outliers=c(length(outliers_bac)))
outliers %>%
  gt() %>%
  #tab_header(title="Number of outliers") %>%
  tab_style(style = cell_text(align="center"), locations = cells_body()) %>%
  tab_style(style = list(cell_text(weight = "bold", align = "center"), cell_fill("grey96")), locations = list(cells_column_labels())) %>%
  tab_options(table.width = px(250))

```

```{r outlier removal}

#bacteria
samples_to_remove_bac <- outliers_bac

DESIGN <- DESIGN[!(DESIGN$ID %in% samples_to_remove_bac),]
COUNT <- COUNT[,!(colnames(COUNT) %in% samples_to_remove_bac)]
ASVs_to_remove <- names(which(rowSums(COUNT) == 0))
COUNT <- COUNT[!(rownames(COUNT) %in% ASVs_to_remove),]
COUNT_norm <- COUNT_norm[,!(colnames(COUNT_norm) %in% samples_to_remove_bac)]
ASVs_to_remove <- names(which(rowSums(COUNT_norm) == 0))
COUNT_norm <- COUNT_norm[!(rownames(COUNT_norm) %in% ASVs_to_remove),]
TAXA <- TAXA[!(rownames(TAXA) %in% ASVs_to_remove),]

```


## Remaining sample size

We end up with the following number of samples per treatment for the analysis.  

```{r sample size rare}

#final number of samples
table_bac <- table(DESIGN$Lake) %>% 
  as.data.frame() %>% 
  pivot_wider(names_from="Var1",values_from="Freq")
#nice output table
gt_nsamples <- function(table, subtitle=subtitle){
  table %>%
  gt() %>%
  #tab_header(title="Sample profile", subtitle = subtitle) %>%
  tab_style(style = cell_text(align="center"), locations = cells_body()) %>%
  tab_style(style = list(cell_text(weight = "bold")), locations = list(cells_group())) %>%
  tab_style(style = list(cell_text(weight = "bold", align = "center"), cell_fill("grey96")),locations = list(cells_column_labels(), cells_column_spanners())) %>%
  tab_style(style = list(cell_text(weight = "bold"), cell_fill("grey98")),locations = list(cells_row_groups())) %>%
  tab_options(table.width = px(250))
}

gt_nsamples(table_bac, subtitle="Bacteria")

```

\pagebreak

## Phytoplankton

We filter and continue only with phytoplankton classes.

### Microbial abundances plots

```{r subset to phytoplankton}

#keep only classes belonging to Phytoplankton
phytoplankton_classes <- c("Cyanophyceae", "Trebouxiophyceae", "Eustigmatophyceae", "Chlorophyceae", 
                           "Bangiophyceae", "Fragilariophyceae", "Coscinodiscophyceae", "Ulvophyceae", 
                           "Compsopogonophyceae", "Cryptophyceae", "Zygnemophyceae", "Bacillariophyceae", 
                           "Florideophyceae", "Dinophyceae", "Chrysophyceae", "Chlorokybophyceae", 
                           "Stylonematophyceae", "Mamiellophyceae", "Dictyochophyceae", "Synurophyceae",
                           "Xanthophyceae", "Mediophyceae", "Prasinodermophyceae", "Chloropicophyceae", 
                           "Phaeophyceae", "Palmophyllophyceae", "Bolidophyceae", "Pedinophyceae", 
                           "Rhodellophyceae", "Glaucocystophyceae", "Euglenida")

to_keep <- which(TAXA$class %in% phytoplankton_classes)
TAXA <- TAXA[to_keep ,]
COUNT <- COUNT[rownames(TAXA),]
COUNT_norm <- COUNT_norm[rownames(TAXA),]

#TSS normalisation
COUNT_norm <- apply(COUNT, 2, function(x) x *100 / sum(x)) #colSums(COUNT_norm) should be 100% per column

```

```{r function to create melted PHYSEQs, add missing samples and plot}

# Data as PHYLOSEQ object

get_PHYSEQ_objects <- function(design, data, taxa){

#change order for melting later and (remove species for fungi and oomycetes)
  taxa <- taxa[,c("kingdom", "phylum", "class", "order", "family", "genus", "species", "ASV_ID")]
  
  #create phyloseq::phyloseq objects
  design <- as.data.frame(design)
  rownames(design) <- design$ID
  data <- as.matrix(data)
  taxa <- as.matrix(taxa[rownames(data),])
  physeq <- phyloseq::phyloseq(sample_data(design),
                     otu_table(data, taxa_are_rows=T),
                     tax_table(taxa))
  
  # agglomerate data by labels, norm and melt
  physeq <- phyloseq::tax_glom(physeq, c("class", "phylum")) #add taxonomy
  physeq <- phyloseq::transform_sample_counts(physeq, function(x) 100 * x/sum(x)) 
  physeq_melt <- phyloseq::psmelt(physeq) 
 
  #define low and high abundant phyla and classes
  threshold <- 1 #%
  
  abu_phyla <- physeq_melt %>% group_by(phylum) %>% summarise(total_abundance_phyla = (sum(Abundance))/(length(unique(Sample))), .groups = "drop") #abu scaled to 100%
  abu_classes <- physeq_melt %>% group_by(class) %>% summarise(total_abundance_classes = (sum(Abundance))/(length(unique(Sample))), .groups = "drop") #abu scaled to 100%
  
  high_abu_phyla <- abu_phyla$phylum[abu_phyla$total_abundance_phyla >= threshold]
  high_abu_classes <- abu_classes$class[abu_classes$total_abundance_classes >= threshold]
  
  #high abundant if phyla and class over 1% of abundance
  physeq_melt$phylum_label <- ifelse(physeq_melt$phylum %in% high_abu_phyla & physeq_melt$class %in% high_abu_classes, physeq_melt$phylum, "Low abundant phyla") #create labels
  physeq_melt$class_label <- ifelse(physeq_melt$phylum %in% high_abu_phyla & physeq_melt$class %in% high_abu_classes, physeq_melt$class, "Low abundant classes") #create labels

  #sort by abundances
  physeq_melt_sorted <- physeq_melt %>%
    left_join(abu_phyla, by = "phylum") %>%
    left_join(abu_classes, by = "class") %>%
    arrange(desc(total_abundance_phyla), desc(total_abundance_classes), desc(Abundance)) %>%
    dplyr::select(!c(total_abundance_phyla, total_abundance_classes))
  
  #make factors
  levels_phyla <- c(setdiff(unique(physeq_melt_sorted$phylum_label), "Low abundant phyla"), "Low abundant phyla") #sorted by abundance with low abu at the end
  physeq_melt_sorted$phylum_label <- factor(physeq_melt_sorted$phylum_label, levels_phyla)
  levels_classes <- c(setdiff(unique(physeq_melt_sorted$class_label), "Low abundant classes"), "Low abundant classes") #sorted by abundance with low abu at the end
  physeq_melt_sorted$class_label <- factor(physeq_melt_sorted$class_label, levels_classes)

  #return(list(physeq, physeq_melt, bcol_class))
  return(list(physeq, physeq_melt_sorted))

}

plot_abundance <- function(physeq_melt, title, phyla_colors){

  #create df for the plot
  df <- physeq_melt[colnames(physeq_melt) %in% c("Sample", "Lake", "phylum_label", "class_label", "Abundance")]
  
  #sort by abundance of most abu phyla
  phyla_abu <- df %>% group_by(phylum_label) %>% summarise(tot_abu=sum(Abundance))
  most_abu_phyla <- as.character(phyla_abu$phylum_label[phyla_abu$tot_abu==max(phyla_abu$tot_abu)][1])
  sample_levels <-  df %>% filter(phylum_label == most_abu_phyla) %>% arrange(desc(Abundance)) %>% select(Sample) %>% unlist()
  df <- df %>% mutate(Sample = factor(Sample,levels=sample_levels))
  
  #ylab and ylim for lines
  ylab <- "Relative abundance [%]"
  ylim <- aggregate(df$Abundance, list(df$Sample), function(x) sum(x, na.rm=T))[,2] %>% max()
  
  #plot
  fig <- ggplot(df, aes(x=Sample, y=Abundance)) +
    geom_bar(aes(fill=phylum_label), stat="identity", show.legend = T)+
    scale_fill_manual(values=phyla_colors)+
    xlab("") +
    ylab(ylab) +
    ggh4x::facet_nested( ~ Lake, scale="free_x", space="free_x", nest_line = element_line())+
    geom_segment(x=-Inf,xend=Inf,y=(ylim*1.05),yend=(ylim*1.05), lwd=1)+ #add lines under WT / bx1
    geom_segment(x=-Inf,xend=Inf,y=ylim,yend=ylim,lwd=0.2, col="black")+ #box around grids
    geom_segment(x=-Inf,xend=Inf,y=0,yend=0,lwd=0.2, col="black")+ #box around grids
    #ggtitle(title)+
    theme_minimal()+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) #axis.text.x=element_blank(), axis.ticks.x=element_blank()

  return(fig)
}

```

```{r create melted PHYSEQs}

#get PHYSEQ
temp <- get_PHYSEQ_objects(design=DESIGN, data=COUNT_norm, taxa=TAXA)
PHYSEQ <- temp[[1]]
PHYSEQ_melt <- temp[[2]]

rm(temp)

```

```{r abundace plots scaled, echo=F, warning=F, message=F}
#| fig-width: 9
#| fig-asp: 0.7 #width to height aspect ratio

#order levels by abu
levels_phyla <- aggregate(PHYSEQ_melt$Abundance, list(PHYSEQ_melt$phylum_label), sum) %>%
  arrange(desc(x)) %>% #sort
  mutate(Group.1 = factor(Group.1, levels = unique(Group.1))) %>% #levels by abu
  mutate(Group.1 = fct_relevel(Group.1, "unassigned", "Low abundant phyla", after = Inf)) %>% #mov unassigned and low abu to the end
  pull(Group.1) %>%
  levels()
  

#define color matrix for phyla
PHYla_colors <- c("#88A0DCFF", "#381A61FF", "#7C4B73FF", "#ED968CFF", "grey70", "grey30")
names(PHYla_colors) <- levels_phyla

rm(temp)


  
PHYSEQ_melt$phylum_label <- factor(PHYSEQ_melt$phylum_label, levels=levels_phyla)

#plot
fig_bac <- plot_abundance(PHYSEQ_melt, "Bacteria: Taxonomy", PHYla_colors)

#combine figures
#combine the plots (library cowplot)
plot_grid(fig_bac + labs(fill="Bacteria: Phylum"),
           align = "hv",
           nrow=1)

```

\pagebreak

## Beta diversity

### PERMANOVA

```{r function for a nice PERMANOVA table}

gt_PERM <- function(dist_paov){
  dist_paov %>%
    gt(rownames_to_stub=T) %>%
    #tab_header(title=title) %>%
    fmt_number(columns = SumOfSqs, decimals=0, sep_mark = "'") %>%
    fmt_number(columns = R2:`Pr(>F)`, decimals=3) %>%
    fmt_missing(missing_text = "-") %>%
    tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = c(`Pr(>F)`),rows = `Pr(>F)` <= 0.5)) %>%
    tab_style(style = cell_text(align="center"), locations = cells_body()) %>%
    tab_style(style = cell_text(size=px(12), align = "center"), locations = cells_title()) %>%
    tab_style(style = list(cell_text(weight = "bold", align = "center"), cell_fill("grey96")), 
              locations = list(cells_stub(),cells_column_labels(), cells_stubhead())) %>%
    tab_options(table.border.top.style = "hidden") %>%
    tab_options(table.width = px(650))
}

```

```{r PERMANOVA}

set.seed(100)
bdist <- vegdist(t(COUNT_norm), method="bray")
bdist_paov <- adonis2(bdist ~ Lake, data=DESIGN, by = "terms")
gt_PERM(bdist_paov)

```

```{r function to plot PCoA Reck}

#function to get centroid lines
centroid_lines <- function(PHYSEQ, dist_matrix, group, axes=c(1,2)){
  #get ordination values
  lines <- as.data.frame(phyloseq::plot_ordination(PHYSEQ, dist_matrix, axes=axes, justDF=T))
  colnames(lines)[1:2] <- c("Axis.1","Axis.2")
  lines <- lines[,colnames(lines) %in% c("Axis.1","Axis.2",group)]#subset
  #get centroids (means of each group)
  means1 <- aggregate(lines$Axis.1, by=as.list(lines[,colnames(lines) %in% group, drop=F]), FUN=mean)
  means2 <- aggregate(lines$Axis.2, by=as.list(lines[,colnames(lines) %in% group, drop=F]), FUN=mean)
  #concat groups (for multiple groups)
  lines$concact <- lines[,colnames(lines) %in% group,drop=F] %>% apply(2, as.character) %>% apply(1, paste, collapse="")
  means1$concact <- means1[,colnames(means1) %in% group,drop=F] %>% apply(2, as.character) %>% apply(1, paste, collapse="")
  means2$concact <- means2[,colnames(means2) %in% group,drop=F] %>% apply(2, as.character) %>% apply(1, paste, collapse="")
  #add means to data.frame
  lines$Axis.1.means <- NA
  lines$Axis.2.means <- NA
  group_values <- unique(lines$concact)
  for (i in group_values) {
    lines$Axis.1.means[lines$concact == i] <- means1$x[means1$concact == i]
    lines$Axis.2.means[lines$concact == i] <- means2$x[means1$concact == i]
  }
  #subset df
  lines <- lines[,colnames(lines) %in% c("Axis.1","Axis.2","Axis.1.means","Axis.2.means")]
  #return
  return(lines)
}

plot_PCoA <- function(PHYSEQ, group, axes=c(1,2), title="", dist_paov, taxa="", n_df_axes=6){
  #ordination
  set.seed(100)
  PCoA_bray <- phyloseq::ordinate(PHYSEQ, method="PCoA", distance="bray")
  p0 <- phyloseq::plot_ordination(PHYSEQ, PCoA_bray, axes = axes)
  #extract data for ggplot
  df <- p0$data
  x_label <- gsub("  ", paste0(taxa, " PCo", axes[1]), p0$labels$x)
  y_label <- gsub("  ", paste0(taxa, " PCo", axes[2]), p0$labels$y)

  #plot
  lines <- centroid_lines(PHYSEQ = PHYSEQ, dist_matrix = PCoA_bray, group = group, axes = axes)
  plot <- ggplot(df, aes_string(x=paste0("Axis.", axes[1]), y=paste0("Axis.", axes[2]), shape="Year", fill="Lake"))+
    geom_segment(aes(xend=lines$Axis.1, yend=lines$Axis.2, color=Lake), x = lines$Axis.1.means, y = lines$Axis.2.means, show.legend = F, lwd=0.3)+ #, color="bisque4"
    geom_point(size=4, stroke=1) +
    scale_fill_manual(name="Lake", values=lake_colors)+
    scale_color_manual(name="Lake", values=lake_colors)+
    scale_shape_manual(name="Lake", values=c("23"=21, "24"=24))+
    theme_bw()+
    theme(legend.position="bottom") + labs(fill="Lake", shape="Year")+
    xlab(x_label)+
    ylab(y_label)+
    ggtitle(title, subtitle = paste0("PERMANOVA\nBray-Curtis. ~ Lake\nLake: R2 = ", 
                                     round(dist_paov$R2[1],3),", p = ",round(dist_paov$`Pr(>F)`[1],3)))+
    scale_x_continuous(labels = scales::number_format(accuracy = 0.01)) + 
    scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
    theme(plot.title = element_text(face="bold"), plot.subtitle = element_text(hjust = 0, size=10))+ #, vjust=-9
    guides(fill=guide_legend(override.aes=list(shape=21), order=3),
           color=guide_legend(override.aes=list(shape=21, fill="grey50"), order=2), shape=guide_legend(order=1))
  
  #expand df for n axes
  for(i in 1:n_df_axes){
    axes <- c(i,i+1)
    set.seed(100)
    PCoA_bray <- phyloseq::ordinate(PHYSEQ, method="PCoA", distance="bray")
    p0 <- phyloseq::plot_ordination(PHYSEQ, PCoA_bray, axes = axes)
    temp <- p0$data %>% 
              dplyr::select(-paste0("Axis.",axes[2])) %>% #rm second axis
              dplyr::relocate(paste0("Axis.",axes[1]), .after = last_col()) #mv axis to the end
    if(i==1){df_return <- temp}else{df_return <- left_join(df_return, temp)} #combine
  }
  
  return(list(plot=plot, df=df_return, PCoA_bray=PCoA_bray))

}

```

```{r PCoA bacteria}
#| fig-width: 8
#| fig-asp: 1.1 #width to height aspect ratio

bPCoA <- plot_PCoA(PHYSEQ, group="Lake", axes=c(1,2), title="PCoA", dist_paov=bdist_paov, taxa="")
bPCoA[["plot"]]

```

```{r save}

#save phytoplankton files
dir.create("../05_OUTPUT")
write.table(COUNT_norm, "../05_OUTPUT/COUNT_phytoplankton.tab", sep="\t")
write.table(TAXA, "../05_OUTPUT/TAXA_phytoplankton.tab", sep="\t")
saveRDS(DESIGN, "../05_OUTPUT/DESIGN.RDS")
saveRDS(PHYla_colors, "../05_OUTPUT/PHYLA_color.RDS")

#save fasta file with phytoplankton species
SEQ <- Biostrings::readDNAStringSet(paths$SEQ)
names(SEQ) <- gsub("OTU", "ASV", names(SEQ))
SEQ <- SEQ[names(SEQ) %in% rownames(TAXA)]
Biostrings::writeXStringSet(SEQ, "../05_OUTPUT/SEQ_phytoplankton.fasta")

```

\pagebreak

## R session information

```{r session info}

sessionInfo()

```

